\documentclass[11pt]{article}
\setlength{\parskip}{0.3em}
\setlength{\textfloatsep}{8pt}
\setlength{\floatsep}{8pt}

\usepackage[a4paper,margin=1in]{geometry}
\usepackage{graphicx}
\usepackage{hyperref}
\usepackage{float}
\usepackage{listings}
\usepackage{xcolor}
\usepackage{tikz}

\usetikzlibrary{arrows.meta, positioning}

\hypersetup{
    colorlinks=true,
    linkcolor=blue,
    urlcolor=blue
}

\title{DnD Campaign Manager API\\
\large COMP3011 Web Services CW1}
\author{Rory Edmonds}
\date{\today}

\begin{document}

\maketitle

\tableofcontents
\newpage


%------------------------------------------------
\section{Introduction}

This report describes the design and implementation of a RESTful web service for managing Dungeons and Dragons campaigns. This system allows Dungeon Masters (DMs) to record campaigns, sessions, encounters, and gameplay events while providing analytics derived from recorded events.

It is implemented using \textbf{FastAPI}, \textbf{SQLAlchemy}, and \textbf{SQLite} and follows a layered architecture separating routing logic, business logic, repositories, validation schemas, and database models.

The system supports both manual gameplay logging and automated encounter simulation, with both providing flexible analytics but only manual gameplay events contributing to campaign-wide analytics.


%------------------------------------------------
\section{System Model}

The API models a Dungeons and Dragons campaign using a hierarchical
structure:

\begin{center}
Campaign $\rightarrow$ Session $\rightarrow$ Encounter $\rightarrow$ Event
\end{center}

\begin{itemize}
\item \textbf{Campaign}: the overall long-running narrative managed by the Dungeon Master.
\item \textbf{Session}: a single play session within a campaign.
\item \textbf{Encounter}: a combat interaction that may be simulated or manually logged.
\item \textbf{Event}: an atomic action such as damage, healing, or spell usage.
\end{itemize}

This structure allows gameplay to be recorded at a granular level while
enabling analytics across encounters, sessions, and entire campaigns.

In addition to the campaign hierarchy, the system models entities that
participate in encounters.

\subsection{Characters and Encounter Participants}

Characters represent persistent entities belonging to a campaign, typically
the player characters. Characters store long-term attributes such as class,
level, maximum hit points, equipment, and known spells.

During combat, characters are instantiated as \textit{encounter participants}.
Participants represent entities that appear in a specific encounter and
track temporary state such as current hit points and remaining spell slots.

Separating characters from encounter participants ensures that encounter
state does not permanently modify the underlying character record.

Participants may also represent monsters imported from external datasets.
This allows encounters to include enemies without adding them to the
campaign's persistent character registry.

\subsection{External Game Data}

The system integrates external datasets describing monsters, spells, and
equipment. These datasets allow the API to provide searchable endpoints
for importing official DnD content into characters and encounters without
requiring the data to be stored directly in the application database.
%------------------------------------------------
\section{System Architecture}

The application follows a layered architecture that separates concerns between HTTP routing, business logic, data access, and persistence. Thhis 
improves maintainability and testability by isolating different aspects of the system. Avoiding tight coupling between layers allows for easier modifications and extensions in the future.

\subsection{Request Processing Flow}
\begin{minipage}{0.48\textwidth}
\centering
\begin{figure}[H]
\centering
\begin{tikzpicture}[
    node distance=1.5cm,
    box/.style={
        rectangle,
        draw,
        rounded corners,
        minimum width=3.2cm,
        minimum height=0.8cm,
        align=center
    }
]

\node[box] (client) {Client};
\node[box, below of=client] (routes) {Routes};
\node[box, below of=routes] (services) {Services};
\node[box, below of=services] (repos) {Repositories};
\node[box, below of=repos] (orm) {ORM};
\node[box, below of=orm] (db) {Database};

\draw[->] (client) -- (routes);
\draw[->] (routes) -- (services);
\draw[->] (services) -- (repos);
\draw[->] (repos) -- (orm);
\draw[->] (orm) -- (db);

\end{tikzpicture}
\caption{Layered architecture of the Campaign Manager API}
\end{figure}
\end{minipage}
\hfill
\begin{minipage}{0.48\textwidth}

When a client sends a request to the API, it is processed through the
architecture layers in sequence.


\begin{enumerate}
\item Client sends HTTP request
\item Route validates input using Pydantic
\item Service layer executes business logic
\item Repository performs database operations
\item SQLAlchemy interacts with SQLite
\item JSON response returned to client
\end{enumerate}
This layered flow ensures that request validation, business logic, and database access remain
clearly separated.
\end{minipage}

\subsection{Architecture Layers}

\begin{itemize}
\item \textbf{Routes}: define HTTP endpoints, validate requests using Pydantic schemas, and return JSON responses. Database sessions are supplied through FastAPI's dependency injection system, ensuring consistent connection handling and allowing dependencies to be overridden during testing.

\item \textbf{Services}: contain business logic such as encounter simulation, event processing, and analytics calculations.

\item \textbf{Repositories}: encapsulate database access and isolate query logic from the rest of the application.

\item \textbf{Models}: define the relational schema using SQLAlchemy ORM.

\item \textbf{Schemas}: provide request validation and response serialisation.
\end{itemize}

%------------------------------------------------
\section{Data Model Design}

Campaign data is stored using a relational database schema implemented
with SQLAlchemy ORM. A relational model was chosen because the system
contains strongly structured relationships between entities such as
campaigns, sessions, encounters, and participants.

\begin{figure}[H]
\centering
\begin{tikzpicture}[
    node distance=1.8cm and 2.5cm,
    entity/.style={
        rectangle,
        draw,
        rounded corners,
        minimum width=4.2cm,
        minimum height=1cm,
        align=center
    }
]

\node[entity] (campaign) {Campaign\\id, name, description};

\node[entity, below=of campaign] (session) {Session\\id, campaign\_id, date};

\node[entity, below=of session] (encounter) {Encounter\\id, session\_id, name};

\node[entity, left=of session] (character) {Character\\id, campaign\_id, name, class};

\node[entity, left=of encounter] (participant) {EncounterParticipant\\id, encounter\_id, character\_id, name};

\node[entity, below=of encounter] (event) {Event\\id, encounter\_id, source\_participant\_id, target\_participant\_id, kind};

\draw[->] (campaign) -- node[right]{0..*} (session);
\draw[->] (session) -- node[right]{0..*} (encounter);
\draw[->] (encounter) -- node[above]{0..*} (participant);
\draw[->] (encounter) -- node[right]{0..*} (event);
\draw[->] (campaign) -- node[above]{0..*} (character);
\draw[->] (character) -- node[right]{0..*} (participant);
\draw[->] (participant) -- node[above]{0..*} (event);

\end{tikzpicture}
\caption{Core Entity Relationships in the Campaign Manager Database}
\end{figure}

The diagram shows the core entity relationships. Campaigns contain
sessions, sessions contain encounters, and encounters contain
participants and events. Characters exist at campaign level as
persistent entities, while encounter participants represent combatants
within a specific encounter and track temporary combat state.

\subsection{Event-Based Logging}

Gameplay interactions are stored as atomic events, such as damage,
healing, or spell usage. Storing events rather than derived statistics
allows analytics to be generated dynamically by aggregating event data.
It also provides a chronological record of how encounters unfold and
allows new analytics features to be added without modifying the schema.

%------------------------------------------------
\subsection{Characters and Encounter Participants}

\textbf{Characters} represent player characters or non-player characters belonging to a campaign. Characters store persistent attributes such as class, level, and equipment.
While \textbf{Participants} represent entities within an encounter. A participant may be linked to a character or may represent a monster imported from an external dataset.
Separating characters from participants allows encounter-specific state such as hit points and spell slots to be tracked independently from persistent character data. It also
allows encounters to include temporary entities such as monsters without cluttering the campaign's character registry.

%------------------------------------------------
\section{Core Features}

The system provides several core features that support gameplay tracking
and analysis.

\textbf{Encounter Simulation} allows encounters to be automatically
resolved. The simulation randomises initiative order, iterates through
combat rounds, selects actions for participants, and generates damage or
healing events using the same event structure as manual gameplay logs.

\textbf{Analytics} are derived by aggregating event records. These include
damage leaderboards, damage taken, healing received, and spell slot usage.

\textbf{External Datasets} are used to provide searchable game content,
including spells, equipment, monsters, and character classes. This allows
official DnD resources to be imported into characters and encounters
without storing the entire dataset in the application database.

%------------------------------------------------
\section{Testing and Deployment}

Automated tests were implemented using \texttt{pytest}. Tests use
FastAPI's \texttt{TestClient} to simulate HTTP requests, allowing the
API to be tested in a way that closely mirrors real usage.

Each test runs against a temporary SQLite database to ensure isolation
from development data and deterministic behaviour. Testing through HTTP
requests validates multiple layers of the application simultaneously,
including request validation, service logic, repository operations, and
database interactions.

Tests verify the behaviour of:

\begin{itemize}
\item campaign CRUD operations
\item session management
\item encounter lifecycle
\item participant creation
\item event logging
\item analytics endpoints
\end{itemize}


The API was developed and tested locally using the FastAPI development
server. At the time of writing the application has not been deployed to
a public environment, although the project structure supports future
deployment using containerisation or cloud hosting platforms such as
Docker, AWS, Railway, or Render.

%------------------------------------------------

\section{Documentation}

The API is documented using FastAPI's automatic OpenAPI generation.
Interactive documentation is available via the Swagger interface during
development, while a static ReDoc export is included with the submission:

\begin{center}
\url{docs/redoc.pdf}
\end{center}

%------------------------------------------------

\section{Limitations and Future Work}

Several limitations exist in the current implementation:

\begin{itemize}
\item encounter simulation uses simplified combat rules
\item authentication and multi-user support are not implemented
\item external datasets are accessed remotely rather than locally cached
\end{itemize}

Future improvements could include richer simulation mechanics, improved
analytics visualisation, user authentication, multi-DM campaign support,
and deployment to cloud infrastructure.

%------------------------------------------------

\section{Conclusion}

This project demonstrates the design and implementation of a RESTful API
for managing tabletop campaign data. The system records gameplay through
event-based logging and generates analytics by aggregating recorded
events.

The layered architecture and modular design allow the system to be
extended with additional gameplay mechanics, analytics capabilities,
and deployment options.

Overall, this project demonstrates how software engineering principles
can be applied to the design of a flexible and extensible web service.

%------------------------------------------------
\section{Powerpoint Presentation}
Here is the link to the PowerPoint presentation for the project:
\begin{center}
    \url{}
\end{center}
%------------------------------------------------
\section{AI Declaration}

Generative AI tools were used during development primarily as a
programming assistant. AI was used to generate boilerplate code, review
architectural decisions, comment + neaten code, suggest testing strategies, and provide
feedback on documentation/report structure and content.

All suggestions were manually reviewed and adapted before being
integrated into the project. Final design decisions, implementation,
and code structure were determined by the author.

\end{document}